HMMotion: Using tracking data to predict coverage strategies in American Football

### 2025-01-05 submission.Rmd
Sub

To accurately forecast the defensive scheme (man- or zone defense) for every play, we need to create various features derived from the tracking data. In particular, we conducted the following feature engineering steps: We first consider all 11 players on each side of the field and compute features related to the convex hull of the positions of the players. In particular, for defense and offense, we compute the area spanned by the convex hull of all player such as well as the largest $y$ distance (i.e.\ the width of the hull) and the largest $x$ distance (i.e. the length of the hull). In addition, we select the five most relevant players on each side of the field. For offense, we omit the offensive line and the QB, while, for defense, we omit nose tackles, defensive tackles and defensive ends, and select the five defenders that were the closest to the five attackers corresponding to a weighted euclidean distance, putting much more emphasis on the y-axis. Finally, we use their standardized $x$ and $y$ coordinates as features and order defensive and offensive players according to their $y$ coordinates. Additionally, for each of the relevant defenders, we compute distances to the football and their orientation with respect to the quarterback (`r colorize("values taken at event lineset. Rouven: Die Values haben wir doch rausgenommen aus dem finalen Modell oder? Robert: ich denke das sollten wir noch besprechen. Ich habe ohnehin mehrere Modelle gefitted...","orange")`). Finally, we extract relevant information from play-by-play data, such as quarter, down, yards to go, home and away score and the remaining game time in the current half (in seconds).

@@

In general, we have a limited dataset available (only 3985 plays) and therefore need to balance the complexity of the model, i.e. we want to control the number of features available. Using the previously described features, we obtain 67 variables basic features from tracking and play by play data, which can be used for all 3 models above. With a more profound dataset, using all of these features (and possibly more) certainly makes sense. However given the small dataset, we focus on only 32 basic features: 6 convex hull related features (3 for offense, 3 for defense), 20 player positions features (10 standardized $x$ and $y$ coordinates, 5 for offense, 5 for defense), and 6 play by play features. In the appendix, we provide more details and analyses using more features. 
Finally we need a suitable model class for predicting man vs zone coverage and chose the following two: On the one hand, we fit a `glmnet` model, which performs implicit feature selection and is able to handle multicollinearity. On the other hand we use an `xgboost` model, which is able to capture non linear effects (and interactions) and in general also handles collinearity, but needs careful hyperparameter tuning and generally performs better on bigger data sets. For all of the model we use careful 10-fold cross validation on a suitable hyperparameter grid.

@@

We compare the predictive performance of our 3 model to determine the effectiveness of player movements before snap for detecting the correct defensive scheme. We train our the models on a 85% of the data and evaluate performance on the remaining 15%. Since we use cross-validation for tuning of the models during the training phase, and this relies on randomly dividing the data into folds, we perform our evaluation repeatedly. We fix 50 different seeds (that is, for tuning, the data is split differently each of the 50 times), and for each we tune and train a model and then evaluate performance of this model on the test set. We do this in order to ensure that our results are meaningful and not due to chance (especially in our small data set).

@@

>>>>

To accurately forecast the defensive scheme (man- or zone defense) for every play, we need to create various features derived from the tracking data. In particular, we conducted the following feature engineering steps: We first consider all 11 players on each side of the field and compute features related to the convex hull of the positions of the players. In particular, for defense and offense, we compute the area spanned by the convex hull of all player such as well as the largest $y$ distance (i.e.\ the width of the hull) and the largest $x$ distance (i.e. the length of the hull). In addition, we select the five most relevant players on each side of the field. For offense, we omit the offensive line and the QB, while, for defense, we omit nose tackles, defensive tackles and defensive ends, and select the five defenders that were the closest to the five attackers corresponding to a weighted euclidean distance, putting much more emphasis on the y-axis. Finally, we use their standardized $x$ and $y$ coordinates as features and order defensive and offensive players according to their $y$ coordinates. Additionally, for each of the relevant defenders, we compute distances to the football and their orientation with respect to the quarterback. Finally, we extract relevant information from play-by-play data, such as quarter, down, yards to go, home and away score and the remaining game time in the current half (in seconds).

@@

In general, we have a limited dataset available (only 3985 plays) and therefore need to balance the complexity of the model, i.e. we want to control the number of features available. Using the previously described features, we obtain 67 variables basic features from tracking and play by play data, which can be used for all 3 models above. With a more profound dataset, using all of these features (and possibly more) certainly makes sense. However given the small dataset, we focus on only 32 basic features: 6 convex hull related features (3 for offense, 3 for defense), 20 player positions features (10 standardized $x$ and $y$ coordinates, 5 for offense, 5 for defense), and 6 play by play features. In the appendix, we provide more details and analyses using more features. 
Finally we need a suitable model class for predicting man vs zone coverage and chose the following two: On the one hand, we fit a `glmnet` (elastic net) model, which performs implicit feature selection and is able to handle multicollinearity. On the other hand we use an `xgboost` model, which is able to capture non linear effects (and interactions) and in general also handles collinearity, but needs careful hyperparameter tuning and generally performs better on bigger data sets. For all of the model we use careful 10-fold cross validation on a suitable hyperparameter grid.

@@

We compare the predictive performance of our 3 models to determine the effectiveness of player movements before snap for detecting the correct defensive scheme. We train our the models on 85% of the data and evaluate performance on the remaining 15%. Since we use cross-validation for tuning of the models during the training phase, and this relies on randomly dividing the data into folds, we perform our evaluation repeatedly. We fix 50 different random seeds (that is, for tuning, the data is split differently each of the 50 times), and for each we tune and train a model and then evaluate performance of this model on the test set. We do this in order to ensure that our results are meaningful and not due to chance (especially in our small data set).

@@

The plots above show the results for our 50 repetitions of the experiment. First, we show the classification accuracy, when using a threshold of 50%. Since this is an arbitrary value and our data is slightly imbalanced (~ 75% zone coverage plays and ~25% man coverage plays), accuracy is not the best metric for analyzing the results. Therefore we also present the AUC values of the model (`r colorize("We also provide further metrics in the Appendix","orange")`). For both metrics, we notice a few thing. On the one hand, the more flexible `xgboost` model performs better than the `glmnet`. However, there is more variation (larger boxes) for the `xgboost` model, suggesting that more data would help improve the tuning of the models. On the other hand, the pre-motion model clearly performs worst. As soon as we incorporate motion information, there is a large increase in performance. While this is expected, we also see that the enhanced model outperforms the model with only naive post-motion information. This shows, that the HMM features capture information, which is distinct from simple motion features and provide another dimension of information, which needs to be incorporated.
